---
title: "Lab 02 - R Essentials"
format:
  html:
    embed-resources: yes
output:
  html_document:
    df_print: paged
---

# Learning goals

Today’s lab is the first steps toward designing a response-adaptive randomization (RAR) trial.  RAR designs are used in precision medicine trials, such as the [BATTLE trial](https://aacrjournals.org/cancerdiscovery/article/1/1/44/2198?casa_token=pK1gZcX-FgkAAAAA:KmsD6qnoaOMxqHJlg0VGlmqr2nqIl49Xupuh0FX7nnJXNjtdBwVWsdmVtUIXKdEWQ_e5i9pG), to gather early evidence of treatment arms that work best for a given biomarker.  Throughout RAR, the treatment allocation adjusts depending on which treatment arm looks most promising.  We will focus on the initial steps of coding this design.

The lab is motivated by the paper by Kurt Viele: [Comparison of methods for control allocation in multiple arm studies using response adaptive randomization](https://journals.sagepub.com/doi/pdf/10.1177/1740774519877836).  It practices:

* Pre-allocating vectors
* Using a loop
* Writing a function

# Notation 

Notation and criteria for study to successfully declare a treatment as efficacious:

* $i = 1, \dots, N$ participants
* $t = 0, 1, 2, 3$ study arms ($t = 0$ is control)
* $Y_i \mid t \sim$ Bern($p_t$) and $y_t$ is a vector of $n_t$ observed outcomes on arm $t$
* The prior on $p_t \sim$ Beta($\alpha_t, \beta_t$)

Posterior Distribution Pr($p_t$ | $y_t$) $\sim$ Beta($\alpha_t + \sum y_t, \beta_t + n_t - \sum y_t$)

Quoting from the paper: The trial is considered successful at the final analysis if there is a high posterior probability that at least one arm has a higher rate than control.

max$_t$ Pr( $p_t > p_0$ ) > $\delta$

where $\delta$ is a threshold chosen to maintain familywise type I error for the study at one-sided 2.5\%.


# Consider 2 different designs:

1.	Equal allocation to four arms throughout design.
2.	RAR where the allocation probability is updated at an interim analysis as follows:

* $V_t = P_t$ (Max)
* $V_0 = min\{\sum V_t \frac{(n_t + 1)}{(n_0 + 1)}, max(V_1, V_2, V_3) \}$

$V_0, V_1, V_2,$ and $V_3$ are renormalized to sum to 1 and are allocation probabilities.

Note: A way to estimate $P_t$(Max) is to `cbind` K = [1000] draws from the posterior distribution of each arm and to see how frequently (across the K draws from each arm) each arm is drawn to be the largest.


# Lab task:

Write a function for each study design to simulate one trial.  

* N = 228 with interim analyses after every 40th participant starting at 40.
* Use equal allocation for first 40 patients for both designs.
* Assume a setting where treatment effect is 0.35 for each study arm (the null scenario).  (But allow flexibility in function for other treatment effects).
* $\alpha_t = 0.35$ for all $t$ and $\beta_t = 0.65$ for all arms.
* Use the following $\delta$ thresholds to determine a successful trial:
  * Design 1, $\delta = 0.9912$
  * Design 2, $\delta = 0.9892$

For simplicity, have your function return a list of at least the following output:

1. The probability that the best treatment arm is better than control.
2. The number of patients assigned to each treatment arm.

# Design 1 : Equal allocation to four arms throughout design.

```{r}


N <- 228 ## subjects

t <- 0:3 ## 4 arms

id <- 1:228

trt <- rep(0:3,228/4)

treat_effect <- 0.35

nt <- N/4

out_0 <- as.numeric(purrr::rbernoulli(N,treat_effect))

dat <- data.frame(id = id,arm = trt,Yi = out_0)


alpha_T <- 0.35

beta_T <- 0.65


delta <- 0.9912
  


# posterior

Design.one <- function(alph, bet, data, N_t,delta){
  
  postr <- matrix(nrow = 1000,ncol = 4)
  
  for(trtt in 1:4){
    dts <- data %>% filter(arm==trtt-1)
    alpha <- alph + sum(dts$Yi)
    beta <- bet + N_t - sum(dts$Yi)
    pt <- rbeta(1000,alpha, beta)
    
    postr[,trtt] <- pt
    
  }
  
  
  
  colnames(postr) <- c(0:3)
  
  Prob <- c("Pr(p_1 > p_0)" = mean(postr[,"1"]>postr[,"0"]),
            "Pr(p_2 > p_0)"= mean(postr[,"2"]>postr[,"0"]),
            "Pr(p_3 > p_0)" = mean(postr[,"3"]>postr[,"0"]))
  
  Design_1_success <- max(Prob) > delta
  
  n_t_s <- rep(N_t,4)
  
  names(n_t_s) <- c(0:3)
  
  
  outs <- list(Probs = Prob, Patients = n_t_s,
               Design.Sucessfull = Design_1_success)
  
  return(outs)
}



Design.one(alph = alpha_T,bet = beta_T,data = dat,N_t = nt, delta = delta)


```



# Design 2 : RAR where the allocation probability is updated at an interim analysis


```{r}


Posterior.draw <- function(alpha_t,beta_t,Y_t,draws) {
  n <- length(Y_t)
  rbeta(draws, alpha_t + sum(Y_t), beta_t + n - sum(Y_t))
}

V_0 <- function(V_t, n_t) {
  min(sum(V_t[-1] * (n_t[-1] + 1) / (rep(n_t[1], 3) + 1)), max(V_t[-1]))
}

RAR <- function(N, seed = 101, draws, alpha_t, beta_t, trt.eff) {
  set.seed(seed)
  
  n_t <- rep(10, 4)
  Y <- matrix(rbinom(sum(n_t), 1, trt.eff), ncol = 4)
  colnames(Y) <- c("0", "1", "2", "3")
  
  P_t <- t(apply(Y, 2, function(y_t) Posterior.draw(alpha_t, beta_t, Y_t = y_t, draws)))
  
  P_t$max <- apply(P_t, 1, which.max)
  V_t <- prop.table(table(P_t$max))
  V_t[1] <- V_0(V_t, n_t)
  V_t <- V_t / sum(V_t)
  
  N <- N - 40
  
  while (N > 0) {
    new.allocate <- sample(0:3, N, prob = V_t)
    n_t <- n_t + table(new.allocate)
    Y_new <- matrix(rbinom(sum(n_t), 1, trt.eff), ncol = 4)
    colnames(Y_new) <- c("0", "1", "2", "3")
    Y <- cbind(Y, Y_new)
    P_t <- t(apply(Y, 2, function(y_t) Posterior.draw(alpha_t, beta_t, Y_t = y_t, draws)))
    P_t$max <- apply(P_t, 1, which.max)
    V_t <- prop.table(table(P_t$max))
    V_t[1] <- V_0(V_t, n_t)
    V_t <- V_t / sum(V_t)
    N <- N - 40
  }
  
  pair.compare <- apply(P_t[, c("1", "2", "3")], 2, function(i) mean(i > P_t$"0"))
  success <- ifelse(max(pair.compare) > 0.5, "Yes", "No")
  success.threshold <- ifelse(max(pair.compare) > 0.9892, "Yes", "No")
  
  list(success = success, success.threshold = success.threshold, 
       ProbBest = max(pair.compare), N = N)
}

#RAR()

```



# If you have more time

* Replicate the design many (10K) times.  Calculate the Type I error.
* Find $\delta$ for each design (supposing you didn't already know it).
* Replicate the study design assuming treatment effects of 
  * $p_0 = p_1 = p_2 = 0.35$
  * $p_3 = 0.65$



